General purpose gene annotation using deep learning

### 2020-01-31 paper/main.tex

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates the generalization capability of the model as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be any of 20.000, 50.000, 100.000 or 200.000 and is determined by the species family(?). Families that tend to have longer genes get assigned longer sequence chunks. If the N75 of a species is less than twice as high as the supposed sequence length it is lowered until either criteria is met or a length of 50.000 is reached. For implementation details on this and the following techniques see Supplementary Material (?) or the source code..

We also construct the final predictions of a single model by overlapping predictions that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to improve model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base.

A model ensemble with 8 components was used to generate the final predictions for each species. For this, we trained 4 seperate animal and plant models and also used a model from an epoch that did not have the highest Genic F1 on the validation set. Depending on the results, we used a model from an epoch with either the highest genic recall or precision. The model with the best Genic F1 was always best in one of those. This was done to increase the diversity of the model ensemble. As with the overlapping, the fusion of the 8 individual predictions was done by a averaging the softmax values of each base pair prediction.

>>>>

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be any of 20.000, 50.000, 100.000 or 200.000 and is determined by the species family(?). Families that tend to have longer genes get assigned longer sequences. If the N75 of a species is less than twice as high as the supposed sequence length it is lowered until either this criteria is met or a length of 50.000 is reached. For implementation details on this and the following techniques see Supplementary Material (?) or the source code..

We also construct the final predictions of a single model by overlapping predictions that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to improve model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. Supplementary Material (?) contains the influence on overlapping with respect to the position in a sequence for each genome, ordered by N75. We found that overlapping works best if the genomes are not very fragmented.

A model ensemble with 8 components was used to generate the final predictions for each species. For this, we trained 4 seperate animal and plant models and also used a model from an epoch that did not have the highest Genic F1 on the validation set. Depending on the results, we used the model iteration with either the highest genic recall or precision. The model with the best Genic F1 was always best in one of those. This was done to increase the diversity of the model ensemble. As with the overlapping, the fusion of the 8 individual predictions was done by averaging the softmax values of each base pair prediction.
